Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-27 analysis/paper/paper.Rmd
add cv test code for metric data to rmd, and revise the para

The development of craft specialization is sometimes associated with social changes since it reflects changes in production systems that could inform on the social, economic, or political conditions of a society. We define specialization as "differential participation in specific economic activities", in which a particular good is produced by fewer producers compared to the total population who consume it [@Costin1991; @Costin2001, pp. 43]. The effects of the emergence of specialization in past societies on their social structures and social processes is an important question in archaeological investigations of small scale societies in the Late Holocene. Considered through a political lens, craft specialization may be a strategy used by elites to control production systems to maintain their power and authority [@Costin2001]. This relationship between craft specialization and social differentiation has been broadly discussed, and many cases show that increasingly specialized production can be a useful indicator of increasing socio-political complexity [@Hirshman2010; @Junker1999]. That said, more economic focus considers production, exchange, and consumption as a whole to explain specialization as part of a strategy to accommodate uneven distributions of resources [@Costin2001; @Arnold1994]. Craft specialisation may appear as a reaction to changes in subsistence practices [@Stark1995economic], or as a result of long-distance trade and inter-regional exchange [@Alizadeh2018]. Present-day investigation of ceramic specialization in the Jodhpur region of India suggests that the adoption of standardized products across cultural boundaries is a reaction to a collapse of previous economic system, indicating a correlation with major socio-economic changes [@Roux2015]. However, it is important to note that explanations for the emergence of specialized production in past societies might not be so straightforward or singular. @Acabado2018's case study in lowland northern Philippines shows that specialization is not necessarily accompanied with social differentiation based on a low degree of ceramic specialization in Ifugao ranked societies during the time of Spanish colonization. The development of specialization could be dependent on unique historical factors, where the conditions and contexts under which specialists organized must be taken into account when making inferences about changes in social organization [@Roux2015; @Costin2001].

The emergence of social inequality in the context of cross-cultural interaction, especially in a colonial situation, has been observed in many parts of world, for example, when foreign trade goods are introduced to local Indigenous societies. The monopolization of long-distance trade goods has caused substantial transformations of Indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005]. Pericolonial archaeology is the study of these indirect effects of colonialism, investigating areas where European colonial rule was limited and their conquests were short-lived and unsuccessful, but their colonial activities had economic and political impacts on Indigenous peoples living on the periphery of colonial control and influence [@Trabert2017; @Acabado2017]. @Acabado2017's study of Ifugao society in the  Philippines highland suggests economic and political intensification during the Spanish presence as the response of Indigenous peoples to the Spanish cooptation. Indigenous societies might not simply passively accept colonial rule, but instead actively negotiate with the colonist, and accommodate or resist foreign intrusion through their daily cultural practices, such as their consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000]. In this paper we present a case study from Northeastern Taiwan as an example of these indirect impacts of foreign occupations on local Indigenous societies. Our aim is to investigate if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in local indigenous societies. Previous reports of archaeological evidence in northeastern Taiwan descrbes increases in the use of prestige goods, such as trade ornaments, and uneven artefact distribution patterns in domestic areas and burial contexts at the time when the Indigenous communities encountered the Spanish and the Dutch in the 17th century. Here we show that in addition to the differential distribution of trade ornaments post-European contact, the locally-made ceramics become highly consistent in their form and shape, compared to other pottery throughout northeastern Taiwan. This shift to higher consistency hints at the possibility of craft specialization in pottery manufacture.

To identify craft specialization in the archaeological record, variation in the standardization of a specialized product is often used as an indicator. This is based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, higher skills, and fewer producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For investigating ceramic standardization, several measurements have been discussed and used, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurements are most widely applied to archaeological assemblages because of the ease of collecting these data. By comparing variations for sets of data using coefficient of variation coupled with statistical testing, the degree of standardization can be identified according to values [@Eerkens2001; @Roux2003; @Stark1995]. @Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be statistically assessed by comparing coefficients of variation of linear dimensions of pots, and differences between artisans can be detected at both intra-individual level and inter-individual level in a region. However, because pottery vessels are typically have curved shapes, linear measurements have limited sensitivity to many kinds of shape variations. Thus, to capture subtle shape variations that might be relevant to standardization, a method that is senstive to non-linear aspects of shape variation is necessary. Here we analyse pottery shape using geometric morphometric methods, a new approach for measuring the degree of standardization because it takes a statistical summary of the overall shape as variable for comparison [@Slice2007]. Using geometric morphometry, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production resulting from cross-cultural interaction contexts.

Our hypothesis is that if foreign influences had impacts on the emergence of social inequality in local Indigenous society due to monopolies of trade arising among a small number of Indigenous people, then we expect to see changes in social organization from a more corporate to a more network organization [@Feinman2000]. We expect this corporate-network shift to manifest as increased specialization pottery manufacture, indicated archaeologically by icreased standardisation of pottery production at Kiwulan. The emergence of craft specialization, here pottery specialization, is often related to changes in social organization towards a society with increasing status inequality. In this case, if the competition among Indigenous individual for foreign resources and trade partnership with European or Chinese colonizers gradually lead to increasing social inequality, then we expect the local ceramics will show more homogeneous shapes after contact due to the emergence of craft specialization caused by rgreater economic and social control by a small group of individuals at Kiwulan. Our results are important for understanding the indirect influences of  European colonists on local Indigenous societies, which remains unclear in East Asia where colonial power had limited success compared to other places in Southeast Asia. Our findings improve our understanding of the relationship between the use of prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers in a pericolonial context [cf. @Acabado2017].

## Geometric Morphometry

Geometric morphometric methods (GMM) has been increasingly applied in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interest. Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to define shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes. There are two common morphometric methods: landmark approaches and outline approaches [@Adams2004]. Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system as xy coordinates for two dimensions, or xyz coordinates for three dimensions. In a typical analysis, starting with raw landmarks, a sample of objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks are minimized using generalized Procrustes analysis (GPA) [@Bookstein1991]. GPA is a method that superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables for further multivariate statistical analyses [@Slice2007]. Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that can provide unambiguous reference points for landmark placement, such as projectile point tips or biological measurements points indicating osteological features [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].

However, one limitation of landmark approaches is that landmarks may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks contains crucial variation. Furthermore, landmark points may be unavailable for structures that are mostly or entirely curves. In those cases, outline approaches are more suitable for analyzing the overall shape of an object. One outline approache is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010]. Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Slice2007; @Gunz2013]. Another approach that is commonly applied to two-dimensional closed shapes is elliptic Fourier Analysis (EFA), that converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients [@Kuhl1982]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Claude2008; @Bonhomme2014]. The number of harmonics determines the quality and precision of the geometric representation of an object. The harmonic power, a cumulated sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014].

EFA is suitable for shapes lacking representative landmarks or where curves contain the most meaningful variation. Applications in archaeology include human remains and zooarchaeology to understand  biological variation, and more recently on stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Ceramics is another archaeological materials that outline approaches can provide new insights into shape variation to answer anthropological questions related to ceramic specialization. Using semi-landmarks method, @Topi2018 examine 89 photographs of globular jars from the Casas Grandes of northwest Mexico. Their results suggest that some ceramic types from the Medio period (AD 1200-1450), were made by specialists based on highly standardized shapes. Among those highly standardized jars, some were made by attached specialists <!-- what does this mean, 'attached' to what? please clarify -->, while others were made by independent specialists according to their spatial distribution patterns. Similarity, @Selden2019 examined 28 Caddo bottle excavated at several sites (AD 500-1700) in northwest Louisiana <!-- using semi-landmarks or something else? please add -->. The results indicate a significant integration between the different parts of the bottles including rim, neck, body, and base that vary in a coordinated manner. Also, two discrete base and body shapes represented northern and southern differences were identified. For the EFA method, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes. Furthermore, @Wilczek201's EFA results help us understand the level of production standardization over time across the region. Those examples show that outline approaches can distinguish variation in ceramic shapes at high resolutions to understand ceramic production.

Taking the ceramics data from Kiwulan, northeastern Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century to get insights into the emergence of ceramic specialization. We use a significance test for the equality of coefficient of variations of shape variables to statistically compare the vessel standarization from different periods. Using pottery shape as a proxy to study craft specialization, we address these questions: Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels? Did pottery shape become more homogeneous and standardized after foreign contacts with European colonizers or Chinese immigrants?

# Archaeological background and materials

Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from the central, undisturbed area of archaeological excavations at Kiwulan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)). Kiwulan is situated on a hill near a riverside at the northern margin of Yilan Province, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides. Yilan is an ideal context to study peripheral colonial influence because Indigenous communities there were isolated by natural barriers that limited the frequency of direct contact with the Spanish and the Dutch settlers in northern Taiwan. The chronology of the archaeological deposits at Kiwulan consists of two cultural components, the upper component and the lower component, with a sterile layer in between [@Chen2007]. This paper focuses on the upper component, dated from AD 1350 to AD 1950, covering the late Iron Age and the historical period, which we define as the European presence in Taiwan in the early 17th century. The Dutch first occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626 <!-- add citation for these dates -->. In 1642, the Spanish was expelled by the Dutch in northern Taiwan, who took over their forts at Helping Dau in Keelung, and in Tamsui. Since then, western Taiwan was mostly under Dutch colonial rule until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].

The archaeological record of Kiwulan's upper component shows traces of foreign contacts, including Europeans in the 17th century, and waves of Chinese immigrants in the 19th century. Imported ceramics from mainland China, stoneware, and ornaments such as beads have been found in the upper component, indicating frequent long-distance trade activities in the 17th century during the European presence in northern Taiwan. In addition to artifacts, archaeological features such as burials, middens, and post-holes with _in-situ_ posts are widespread across  the 1-2 m thick deposit of the upper component, and demonstrate that Kiwulan was a continuously occupied large settlement site [@Chen2007]. To compare different foreign influences, we classified the upper component into three chronological phases: the pre-European, European, and Chinese periods. We identified these phases using 32 previously published radiocarbon ages <!-- add citaion to report with dates -->, excavation depth measurements, stratigraphic details reported by the excavators (colour, disturbance, etc.), and finds of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles used by the Chinese in the 19th century [@Hsieh2009; @Wang2011]. The deposit shows signs of continuous human occupation during each of the three phases across the sampling area.

The most abundant artefacts in the upper component are locally made ceramics, which are distributed throughout the sequence and across the site. More than 550,000 sherds were excavated, and around 1,200 vessels could be reconstructed. There are only two forms of local vessels, a cooking pot and a steamer made of two cooking pots stacked together with a clay filter layer in between. Those pots forms show high consistency in shape. They have a globular body with a short neck and wide mouth. The exterior surface below the neck is decorated with a wide variety of impressed geometric motifs. These vessels were probably used for cooking, with evidence of charred residues and carbon deposits frequently observed on vessel interiors and soot on their exteriors. Moreover, a lack of evidence of other utilitarian earthenware ceramics suggests these globular pots were mainly used for daily cooking. They are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes. Finger impressions and seams usually on the interior indicate they were pinched using slabs of clay and shaped by hand. This kind of pot has been widely found at archaeological sites during the late Iron Age and the historical period throughout the Yilan Plain.

@@

Petrographic analysis for 34 thin sections shows high percentage of inclusions (15-50%), including argillite (15-40%), metasandstone (1-10%),  sandstone fragments (1-6%), quartz (1-5%), and trace amounts of feldspar and slate. The size of the particles ranges from 500 micron to 1300 micron. In general, the globular vessel farbric presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz. The composition matches the mineralogical composition of local raw materials in the Yilan Plain [@Chen2016]. There is no substantial changes in the inclusions over time, indicating continuity in pottery fabric composition across the three periods.

@@

A total of `r total_no` largely intact vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for ellipic Fourier analysis. Most of them cover every vessel component from rim to bottom. To fully explore the pottery standardization, we also take the metric measurements into account for comparison. The total of 362 reconstructed pots (pre-European = 153, European = 173, Chinese = 36) were measured. The metric attributes include thickness of rim, neck, and body, and diameter of rim, neck, and body. For each pot, two measurements for thickness and diameter were took to compute an average value. To obtain information about the pottery shape, we also compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter.

We used scanned pottery drawings acquired from Bureau of Cultural Affairs in Yilan. All drawings present a two-dimensional view of vessel cross-sections with indications of metric measurements. The scanned drawings were imported into Inkscape (http://inkscape.org) where outlines were traced manually. Where only half cross-section images were available, these were duplicated, flipped, and then joined with another one to create a 2D closed outline for each vessel. Geometric morphometric analyses were conducted using the R software [@Rlanguage2019] using the functions included in the Momocs, a R package intended to quantify and analyse shapes, especially for outline analysis [@Bonhomme2014]. The digitised outlines were converted into a list of successive x-y pixel coordinates for elliptic Fourier analysis (EFA), to investigate morphological differences among pottery shapes from our three occupation phases. The harmonic coefficients generated by EFA were analysed by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation through .

## Statistical analysis

The principal component (PC) scores were analysed with a multivariate analysis of variance (MANOVA) test to identify significant differences in shapes between occupation phases. We also computed coefficients of variation values (CVs) for the PCs, treating the PCs as shape variables that are more informative than linear dimensions. The coefficient of variation is a common and widely-used statistical measure of the spread of a set of measurements of a sample. It is defined as the standard deviation divided by the mean in a ratio scale format:

By standardizing standard deviation in each data set, the coefficients of variation values allows us to directly compare variation in samples measured with different units or means. We take this as our measurement of standardisation in vessel shape: lower CV values reflect higher standardisation, and thus increased craft specialisation in the community. To answer the question of whether CV values for vessels samples across our three occupational phases are significantly different or not, we used the "Asymptotic test for the equality of coefficients of variation from k populations" [@Feltz1996], a significance test for CVs provided by the R package cvequality [@Marwick2019]. Our implementation is based on the following from @Feltz1996:

$$\widehat{\tau }_c = \frac{\left( \sum _{j=1}^km_j\frac{S_j}{\bar{x}_j} \right)}{M}$$ 
where 
$$M=\sum _{j=1}^m m_j$$

Then the test statistic can be computed with:

$$\begin{aligned} D'A D= \widehat{\tau }_c^{-2}\left( 0.5+\widehat{\tau }_c^2\right) ^{-1}\left[ \sum _{i=1}^km_i\left(\frac{s_i}{\bar{x}_i} \right)^2-\frac{\widehat{\tau }_c^2}{M} \right] \end{aligned}$$

The $D'AD$ value measures how far each sample $c_v$ is from our estimate of the population $c_v$. @Feltz1996 note that the $D'AD$ value distributes as a central $\chi^2$ random variable with $k-1$ degrees of freedom, from which a p-value can be computed.

To enable re-use of materials and improve reproducibility and transparency [@Marwick2017], the entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is openly available online at https://doi.org/10.17605/OSF.IO/ABVGF. Also in this version-controlled compendium [@Marwick2018] are the raw data for all the visualizations and tests reported here. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.

To investigate the effect of the historically-documented decline of Indigenous population in the Chinese period <!-- add citation on population change during Chinese period --> on pottery production we examined the spatial distribution of pots in our sample in each period. Our concern is that the smaller population size during the Chinese period could confound our investigations into craft specialisation beause of reduced numbers of potters and pottery production shrinking a a few locations in the settlement during this time. Altough the overall number of vessels is smaller during the Chinese contact period, Figure \@ref(fig:pottery-spatial-pattern) shows that pottery is distributed randomly across the sampling area without any distinctive clusters. This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialsation.

@@

<!-- basic description of shape -->

>>>>

The development of craft specialization is usually associated with the social changes since it reflects production systems that could inform social, economic, or political conditions of a society. Specialization is defined as "differential participation in specific economic activities", in which a particular good is produced by fewer producers compared to the total population who consume it [@Costin1991; @Costin2001, pp. 43]. The implication of the emergence of specialization for social structure or social process has been a theme in archaeological investigations. From the perspective of the political approaches, craft specialization specifically attached production, is a strategy used by elites to control over production systems to maintain power and authority [@Costin2001]. The relationship between craft specialization and social differentiation is broadly discussed, and many cases show that specialized production is a useful indicator of increasing socio-political complexity [@Hirshman2010; @Junker1999]. One the other hand, economic models consider production, exchange, and consumption as a whole to explain specialization as a subsistence strategy to accommodate uneven distribute of resources [@Costin2001; @Arnold1994], a reaction to the changes in subsistence practices [@Stark1995economic], or an outcome result from by long-distance trade and inter-regional exchange [@Alizadeh2018]. An present-day investigation of ceramic specialization in the Jodhpur region of India suggests that the adoption of standardized products across cultural boundaries is a reaction to a collapse of previous economic system, indicating the correlation with major socio-economic changes [@Roux2015]. However, it is important to note that the explanation for the emergence of specialized production might not be straightforward or exclusive. @Acabado2018's case study in lowland northern Philippines shows that specialization is not necessarily accompanied with social differentiation based on the result of a low degree of ceramic specialization in Ifugao ranked societies during the Spanish colonization. The development of specialization could be history dependent, where condition and contexts under which specialists organized are important to make an inference for changes in social organization [@Roux2015; @Costin2001].

The emergence of social inequality in the context of cross-cultural interaction especially in a colonial situation was observed in many parts of world when the introduction of foreign trade goods to local indigenous societies. The monopolization of long-distance trade goods usually cause substantial transformations of indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005]. Recently indirect effects of colonialism or pericolonial archaeology were address when investigating areas where European colonial rule was limited or their conquests were unsuccessful, but their colonial activities had economically and politically impacts on indigenous peoples in the periphery of colonial control [@Trabert2017; @Acabado2017]. For example, @Acabado2017's study on Ifugao society in the highland Philippines suggests economic and political intensification during the Spanish presence as the response of indigenous peoples to the Spanish cooptation. Indigenous societies might not just passively accept the colonial rules, but instead actively negotiated with the colonist, and accommodate or resist the foreign intrusion through their daily cultural practices, such as consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000]. Northeastern Taiwan is an ideal example to explore the indirect impacts of the foreign occupations on local indigenous societies due to its isolated location.  The archaeological evidence in northeastern Taiwan show evidence of increasing use of prestige goods such as trade ornaments and uneven distribution pattern in domestic area and burial contexts when the society encountered by the Spanish and the Dutch in the 17th century. In addition to the differential distribution of trade ornaments, the locally made ceramic present a high consistency in form and shape compared to other pottery throughout northeastern Taiwan that hints the possibility of pottery specialization.

To identify craft specialization, the variation of standardization of specialized product is commonly used as an indicator based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For ceramic standardization, several measurements have been proposed and discussed, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurement is widely applied to archaeological assemblages. By comparing variations for sets of data using coefficient of variation coupled with statistical testing, the degree of standardization can be identified according to values [@Eerkens2001; @Roux2003; @Stark1995]. @Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be assessed by comparing coefficient of variation and the differences between artisans can be detected at both intra-individual level and inter-individual level in a region. However, linear measurements might be limited for identifying differences in variations because they could be insensitive to subtle variations resulting from changes in assemblages. Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007]. Using geometric morphometrics approaches, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts. The aim of this paper is to investigate if there is increasing ceramic specialization resulting from interaction with the Europeans in the 17th century or the Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in local indigenous societies.

Our hypothesis is that if foreign influences had impacts on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then we expect to see the changes in social organization from a more corporate to a more network organization that would be reflected by pottery production at Kiwulan [@Feinman2000]. The emergence of craft specialization, here pottery specialization, are usually related to changes in social organization towards a society with increasing status inequality. In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to increasing social inequality, then we expect the local ceramic will show more homogeneous features after contact due to the emergence of craft specialization caused by control of small group of individuals. This study is important to understand the indirect influences of the European colonists on the local indigenous societies, which remains unclear in East Asia where the colonial power was not successful compared to other places in Southeast Asia. Also, it would help to understand the relationship among the use of prestige goods, the degree of ceramic specialization, and the influence of foreign colonizers in a pericolonial context [cf. @Acabado2017].

## Geometric Morphometric approaches

Geometric morphometric methods (GMM) has been increasingly applied in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests. Different from linear measurement approaches that capture shapes by length, width or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values of coordinate axes. According to different focuses and nature of original shape data, there are two common morphometric methods, landmark approaches and outline approaches [@Adams2004]. Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system as x, y coordinates for two dimensions, or x, y, z coordinates for three dimensions. Based on distances between raw landmarks, objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks are minimized using generalized Procrustes analysis (GPA) [@Bookstein1991]. GPA is a method that superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables for further multivariate statistics analyses [@Slice2007]. Landmark-based morphometrics are widely applied to archaeological objects with a obvious morphological signature that can provide ideal reference points for landmark placement, such as projectile point tips or biologically measurements points indicating osteological features [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].

However, one limitation of landmark approaches is that landmarks may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks are crucial for variation. Furthermore, representative and reliable landmark points may be unavailable for complex or rounded shaped structures. In those cases, outline approaches are better way for analyzing the overall shape of an object. One of the outline approaches is the semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010]. Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Slice2007; @Gunz2013]. Another approach is elliptic Fourier Analysis (EFA) that turns coordinates into coefficients using mathematical function, commonly applied to two-dimensional closed shape [@Kuhl1982]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions, called harmonic coefficients [@Adams2004; @Claude2008; @Bonhomme2014]. The number of harmonics determines the quality and precision of the geometry of an object. The harmonic power, a cumulated sum of squared harmonic coefficient, provides a robust rule for the desired number of harmonics [@Bonhomme2014].

Outline approaches especially EFA is suitable for shapes lacking representative landmarks or curves convey more meaningful variations. The applications in archaeology are seen on human remains or zooarchaeology to understand the biological variation and recently more on stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Ceramics is another archaeological materials that outline approaches can provide new insights into shape variation to answer anthropological questions related to ceramic specialization. Using semi-landmarks method, @Topi2018 examine 89 photographs of globular jars from the Casas Grandes of northwest Mexico. Their results suggest that some ceramic types from Medio period (AD 1200-1450), were made by specialists based on high standardized shape. Among those high standardized jars, some were made by attached specialists, while others were made by independent specialists according to their spatial distribution patterns. Similarity, @Selden2019 examines 28 Caddo bottle excavated at several sites (AD 500-1700) in northwest Louisiana from the Clarence H. Webb collections. The results indicate a significant integration between the different parts of the bottles including rim, neck, body, and base that vary in a coordinated manner. Also, two discrete base and body shapes represented north and south difference were identified. For the EFA method, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. The results show the variation indicated by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic shapes. Furthermore, the EFA results help to understand the level of production standardization over time across the region. Those examples show that outline approaches can distinguish variation in ceramic shapes in a finer resolution to understand deeply the ceramic production.

Taking the ceramics data from Kiwulan, northeaster Taiwan, we use EFA to evaluate the level of standardization of ceramics in relation to the European presence in the 17th century that might hint the emergence of ceramic specialization. In addition, we use significance test for the equality of coefficient of variation of shape variables to compare the vessel standarization from different periods in a statistical way. Using pottery shape as a proxy to study craft specialization, we answer the questions: Did foreign influence have impacts on indigenous pottery production that can be detected in the shape of the vessels? How does pottery change in shape in relation to foreign influences? Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrants? Starting from the historical background of colonial influences and its reflection in the material record, this study explore the relationship between the emergence of social inequality and craft specialization that can extend our understanding of the reactions of indigenous societies in the periphery of colonial control, providing a perspective that could be applied in other parts of the world in a similar context.

# Materials and archaeological background

Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from the largest excavation section at Kiwulan, northeastern Taiwan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)). Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides . Yilan is an ideal context to study peripheral colonial influence because it was relatively isolated by natural barrier that reduced frequency of direct contact with the Europeans presences, the Spanish and the Dutch settled in northern Taiwan. The chronology of Kiwulan consists of two cultural components, the upper component and the lower component, with a sterile layer in between [@Chen2007]. This paper focuses on the upper component dated from AD 1350 to AD 1950, covering the late Iron Age and the historical period defined by the European presence in Taiwan in the early 17th century. The Dutch firstly came and occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626. In 1642, the Spanish was expelled by the Dutch, who took over their forts at Helping dau in Keelung, and in Tamsui. Since then, western Taiwan was mostly under the colonial rule of the Dutch until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].

The upper component experienced foreign contacts including the Europeans in the 17th century and great waves of Chinese immigrants in the 19th century. Imported ceramics from mainland China, stonewares, and ornaments elements such as beads were trade goods commonly found in the upper component that indicates the frequent long-distance trade activities in the 17th century during the European presence in northern Taiwan. In addition to artifacts, features such as burials, middens, and post-holes with in-situ posts widespread across 1-2 m thick deposit explain that Kiwulan was a continuously occupied and large settlement site [@Chen2007]. To compare different foreign influences, the upper component were classified into three phases, the pre-European, European, and Chinese periods. The principle of identification of three phases is based on 32 radiocarbon ages, excavation depth, consistency of contexts, and types of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars prevailing in the 17th century, and bricks and tiles used by the Chinese in the 19th century [@Hsieh2009; @Wang2011]. The deposit shows signs of continuous human occupation during each of the three phases across the sampling area.

The predominant artefact found is locally made ceramics, which distributed throughout the sequence across the site. More than 550 thousands of sherds were excavated, and around 1,200 vessels were reconstructed. It is interesting to note that there are only two forms of local vessels, one is cooking pot and another one is steamer made of two cooking pots stacked together with a filter clay layer in between. Those pots show high consistency in shape. They are globular body with short neck and wide mouth. The exterior surface below the neck is treated with a wide variety of impressed geometric motifs. They were believed used for cooking based on the evidence of charred residues and carbon deposits commonly observed on the interior and soot on the exterior. Moreover, a lack of evidence of other utilitarian earthenware ceramics suggests this globular pot was mainly used for daily cooking. They are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes. Finger impressions and seams usually on the interior indicate they were pinched using slabs of clay and shaped by hand. In addition to Kiwulan site, this kind of pot was widely found at other archaeological sites during the same period in Yilan Plain.

@@

Petrographic analysis for 34 thin sections shows high percentage of inclusions (15-50%), including argillite, quartz, feldspar, metasandstone, sandstone fragment, and slate. The most common type of sherds, also the dominant one, has a high percentage of inclusion (25-40%), which is composed mainly by argillite (15-40%), followed by metasandstone (1-10%), sandstone (1-6%), and quartz (1-5%). The size of the particles ranges from 500 micron to 1300 micron (=1 mm). In general, this type presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz. The composition matches the mineralogical composition of local raw materials in the Yilan Plain [@Chen2016]. It seems there is no obvious changes in the inclusions over time.

@@

A total of `r total_no` relativity intact vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for Ellipic Fourier analysis. Most of them cover every components from rim to bottom. Given that the decline of indigenous population in the Chinese period might have impact on the interpretation, we examined the spatial distribution of pots analyzed in this paper. The pottery distributed randomly across the sampling area without any clusters could rule out the bias caused by the population dynamics (Figure \@ref(fig:pottery-spatial-pattern)). To fully explore the pottery standardization, we also take the metric measurements into account for comparison. The total of 362 reconstructed pots (pre-European = 153, European = 173, Chinese = 36) were measured. The metric attributes include thickness of rim, neck, and body, and diameter of rim, neck, and body. For each pot, two measurements for thickness and diameter were took to compute an average value. To obtain information about the pottery shape, we also compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter.

@@

The elliptic Fourier coefficients of `r total_no` pottery from three phases were calculated. Reliable pottery outline was captured by 13 harmonics that gather 99 % of the total harmonic power. The average shape of vessels from each phase was visualized by computing the mean of the standardized Fourier coefficients within each phase group (Figure \@ref(fig:EFA-analysis)). Figure \@ref(fig:thin-plate-splines) shows the shape changes described using thin-plate spline warping for paired periods, pre and post European periods, and post-European and Chinese period in a time sequence.
